---
title: "Gene Expression Analysis with `myTAI`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene Expression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Introduction

In the [Introduction](https://github.com/HajkD/myTAI/blob/master/vignettes/Introduction.Rmd) vignette we introduced and discussed how phylotranscriptomics can be applied
to capture evolutionary signals in developmental transcriptomes. Furthermore,
in the [Enrichment Analyses](https://github.com/HajkD/myTAI/blob/master/vignettes/Enrichment.Rmd) vignette we provide a use case to correlate specific
groups or sets of genes with their predicted evolutionary origin. Here, we aim to
combine previously introduced techniques with _classic_ gene expression analyses to detect possible functional causes for the observed transcriptome conservation. 

In other words, phylotranscriptomics allows us to detect stages or periods of evolutionary conservation and is able to predict the evolutionary origin of process or trait specific genes
based on enrichment analyses. By combining evolutionary enrichment analyses with the
functional annotation of process or trait specific genes (see [Functional Annotation](https://github.com/HajkD/biomartr/blob/master/vignettes/Functional_Annotation.Rmd) and [Phylotranscriptomics](https://github.com/HajkD/biomartr/blob/master/vignettes/Phylotranscriptomics.Rmd) for details) the detection of evolutionary signals can be correlated with 
functional processes. Then, performing gene expression analyses on corresponding process or trait specific genes allows users to detect potential causes of stage/period specific evolutionary transcriptome conservation. 


The following sections introduce main gene expression data analysis techniques implemented in `myTAI`:

- Detection of Differentially Expressed Genes (DEGs) 

- Collapsing Replicate Samples

- Filter for Expressed Genes

- Compute the Statistical Significance of Each Replicate Combination

## Detection of Differenentially Expressed Genes (DEGs) 

A variety of methods have been published to detect differentially expressed genes. Some methods
are based on non-statistical quantification of expression differences (e.g. fold-change and log-fold-change), but most methods are based on statistical tests to quantify the significance 
of differences in gene expression between samples. These statistical methods can furthermore, be divided into two methodological categories: parametric tests and non-parametric tests.
The `DiffGenes()` function available in `myTAI` implements the most popular and useful methods 
to detect differentially expressed genes. In the literature different methods have been introduced
and discussed for microarray technologies versus RNA-Seq technologies. 

In this section we will introduce all methods implemented in `DiffGenes()` using small examples
and will furthermore, discuss published advantages and disadvantages of each method and each mRNA
quantification technology.

## Fold-Changes

A fold change in gene expression is simply the ratio of the gene expression level of one sample against a second sample: $\frac{e_{i1}}{e_{i2}}$, where $e_{i1}$ is the expression level of gene $i$ in sample one and $e_{i2}$ is the expression level of gene $i$ in sample two. In case replicate 
expression levels are present for each sample the ratio of arithmetic means of the corresponding replicates is computed: $\frac{\bar{e}_{i1}}{\bar{e}_{i2}}$, where $\bar{e}_{i1}$ is the arithmetic mean of replicate expression levels of gene $i$ in sample one and $\bar{e}_{i2}$ is the arithmetic mean of replicate expression levels of gene $i$ in sample two.

* __Advantages:__

* __Disadvantages:__


### Example: Fold-Change

```{r, eval=FALSE}
data("PhyloExpressionSetExample")

# Detection of DEGs using the fold-change measure
DEGs <- DiffGenes(ExpressionSet = PhyloExpressionSetExample[1:5,1:8],
                  nrep          = 2,
                  method        = "foldchange",
                  stage.names   = c("S1","S2","S3"))


head(DEGs)
```

```
   Phylostratum      GeneID    S1->S2    S1->S3    S2->S1    S2->S3    S3->S1    S3->S2
1            1 at1g01040.2 1.6713881 2.0806706 0.5983051 1.2448758 0.4806143 0.8032930
2            1 at1g01050.1 1.0273222 1.2709185 0.9734045 1.2371177 0.7868325 0.8083305
3            1 at1g01070.1 1.3087379 1.4044799 0.7640949 1.0731560 0.7120073 0.9318310
4            1 at1g01080.2 0.7779572 0.7286769 1.2854177 0.9366542 1.3723503 1.0676299
5            1 at1g01090.1 0.3803866 0.2288961 2.6289042 0.6017460 4.3687939 1.6618307
```


## Collapsing Replicate Samples

After performing differential gene expression analyses, replicate expression levels
are collapsed to a single stage specific expression level. For this purpose, `myTAI`
implements the `CollapseReplicates()` function, allowing users to combine replicate
expression levels stored in a standard `PhyloExpressionSet` or `DivergenceExpressionSet` object to a stage specific expression level using a specified [window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html).


```{r,eval=FALSE}
library(myTAI)

# load example data
data(PhyloExpressionSetExample)

# genrate an example PhyloExpressionSet with replicates
ExampleReplicateExpressionSet <- PhyloExpressionSetExample[ ,1:8]

# rename stages
names(ExampleReplicateExpressionSet)[3:8] <- c("Stage_1_Repl_1","Stage_1_Repl_2",
                                               "Stage_2_Repl_1","Stage_2_Repl_2",
                                               "Stage_3_Repl_1","Stage_3_Repl_2")
# have a look at the example dataset
head(ExampleReplicateExpressionSet, 5)
```

```
  Phylostratum      GeneID Stage_1_Repl_1 Stage_1_Repl_2 Stage_2_Repl_1 Stage_2_Repl_2
1            1 at1g01040.2       2173.635      1911.2001       1152.555      1291.4224
2            1 at1g01050.1       1501.014      1817.3086       1665.309      1564.7612
3            1 at1g01070.1       1212.793      1233.0023        939.200       929.6195
4            1 at1g01080.2       1016.920       936.3837       1181.338      1329.4734
5            1 at1g01090.1      11424.567     16778.1685      34366.649     39775.6405
  Stage_3_Repl_1 Stage_3_Repl_2
1       1000.253       962.9772
2       1496.321      1114.6435
3        864.218       877.2060
4       1392.643      1287.9746
5      56231.569     66980.3673
```

Now, assume that this example `PhyloExpressionSet` stores three developmental stages and 2
biological replicates for each developmental stage. Of course, we could now compute and visualize the TAI profile by typing:

```{r,eval=FALSE}
# visualize the TAI profile over 3 stages of development
# and 2 replicates per stage
PlotPattern(ExpressionSet = ExampleReplicateExpressionSet,
            type          = "l",
            lwd           = 6)

```

Usually, one would expect that variations in replicate values are smaller than variations between developmental stages. In this example however, we constructed replicate values that
vary larger than expression levels between developmental stages. 
For many applications it might be useful to visualize TAI/TDI values of replicates as well,
but normally replicate values are collapsed to one gene and stage specific value after
differential gene expression analyses and replicate quality control have been performed.

The following example illustrates how to collapse replicates with `CollapseReplicates()`:

```{r,eval=FALSE}
# combine the expression levels of the 2 replicates (const) per stage
# using geom.mean as window function and rename new stages to: "S1","S2","S3"
CollapssedPhyloExpressionSet <- CollapseReplicates(
                                       ExpressionSet = ExampleReplicateExpressionSet,
                                       nrep          = 2,
                                       FUN           = geom.mean,
                                       stage.names   = c("S1","S2","S3"))

# have a look at the collpased PhyloExpressionSet
head(CollapssedPhyloExpressionSet)

```

```
   Phylostratum      GeneID         S1         S2         S3
1            1 at1g01040.2  2038.1982  1220.0147   981.4381
2            1 at1g01050.1  1651.6070  1614.2524  1291.4582
3            1 at1g01070.1  1222.8557   934.3975   870.6878
4            1 at1g01080.2   975.8215  1253.2189  1339.2866
5            1 at1g01090.1 13844.9740 36972.3612 61371.0937
6            1 at1g01120.1   815.3288   894.8987   905.8272
```

The `nrep` argument specifies either a constant number of replicates per stage or
a numeric vector storing variable numbers of replicates for each developmental stage.
In our example, each developmental stage had a constant (equal) number of replicates 
per developmental stage (`nrep = 2`). In case a variable stage specific number of replicates
is present, one could specify  `nrep = c(2,3,2)` defining the case that developmental stage 1 stores 2 replicates, stage 2 stores 3 replicates, and stage 3 again, stores 2 replicates.

The argument `FUN` specifies the window function to collapse replicate expression levels to
a single stage specific value. In this example, we chose the `geom.mean()` (geometric mean) function
implemented in `myTAI`, because our example `PhyloExpressionSet` stores absolute expression levels. Notice that the mathematical equivalent of performing arithmetic mean (`mean()`)
computations on `log` expression levels is to perform the geometric mean (`geom.mean()`) on
absolute expression levels.

The `stage.names` argument then specifies the new names of collapsed stages.


## Filter for Expressed Genes

After differential gene expression analyses and replicate aggregation have been performed,
some studies filter gene expression levels in RNA-Seq count tables or microarray expression matrices for non-expressed genes.
In other words, in most studies performing RNA-Seq experiments for example, FPKM/RPKM values < 1
are remove from the processed (final) count table.

For this purpose `myTAI` implements the `Expressed()` function to filter (remove) expression levels in RNA-Seq count tables which do not pass a defined expression threshold.

The `Expressed()` function takes a standard `PhyloExpressionSet` or `DivergenceExpressionSet` object storing a RNA-Seq count table (CT) or microarray gene expression matrix and removes genes from this count table or gene expression matrix that have an expression level below a defined `cut.off` value.

`Expressed()` allows users to choose from several gene extraction methods (see `?Expressed` for details):

* `const`: all genes that have at least one stage that undercuts the expression `cut.off` will be excluded from the `ExpressionSet`

* `min-set`: genes that undercut the expression level `cut.off` in `ceiling(n/2)` stages will be excluded from the `ExpressionSet`

* `n-set`: genes that undercut the expression level `cut.off` in `n` stages will be excluded from the `ExpressionSet`


```{r,eval = FALSE}

# check number of genes in PhyloExpressionSetExample
nrow(PhyloExpressionSetExample)
#> [1] 25260

# remove genes that have an expression level below 8000
# in at least one developmental stage
FilterConst <- Expressed(ExpressionSet = PhyloExpressionSetExample,
                         cut.off       = 8000,
                         method        = "const")

nrow(FilterConst) # check number of retained genes
#> [1] 449
```

As you can see, only 449 out of 25260 genes in `PhyloExpressionSetExample` have an absolute expression level above `8000` when omitting genes using `method = 'const'`.


## Compute the Statistical Significance of Each Replicate Combination

In some cases (high variability of replicates) it might be useful to verify that there is no sequence of replicates (for all possible combination of replicates) that results in a non-significant `TAI` or `TDI` pattern, when the initial pattern with combined replicates was shown to be significant.

The `CombinatorialSignificance()` function implemented in `myTAI` allows you to compute the p-values quantifying the statistical significance of the underlying pattern for all combinations of replicates.


### A small Example:

Assume a `PhyloExpressionSet` stores 3 developmental stages with 3 replicates measured for each stage. The 9 replicates in total are denoted as: $1.1, 1.2, 1.3, 2.1, 2.2, 2.3, 3.1, 3.2, 3.3$. Now the function computes the statistical significance of each pattern derived by the corresponding combination of replicates, e.g.

- 1.1, 2.1, 3.1 : p-value for combination 1

- 1.1, 2.2, 3.1 : p-value for combination 2

- 1.1, 2.3, 3.1 : p-value for combination 3

- 1.2, 2.1, 3.1 : p-value for combination 4

- 1.2, 2.1, 3.1 : p-value for combination 5

- 1.2, 2.1, 3.1 : p-value for combination 6

- 1.3, 2.1, 3.1 : p-value for combination 7

- 1.3, 2.2, 3.1 : p-value for combination 8

- 1.3, 2.3, 3.1 : p-value for combination 9

- ...

This procedure yields 27 p-values for the $3^3$ ($n^m$) replicate combinations, where
$n$ denotes the number of developmental stages and $m$ denotes the number of replicates per stage.

Note, that in case you have a large amount of stages/experiments and a large amount of replicates the computation time will increase by $n^m$. For 11 stages and 4 replicates, $4^{11}$ = 4194304 p-values have to be computed. Each p-value computation itself is based on a permutation test running with $1,000, 10,000, ...$ or more permutations. Be aware that this might take some time.

The p-value vector returned by this function can then be used to plot the p-values to see whether an critical value $\alpha$ is exceeded or not (e.g. $\alpha = 0.05$).

```{r,eval=FALSE}

# load a standard PhyloExpressionSet
data(PhyloExpressionSetExample)

# we assume that the PhyloExpressionSetExample
# consists of 3 developmental stages
# and 2 replicates for stage 1, 3 replicates for stage 2,
# and 2 replicates for stage 3
# FOR REAL ANALYSES PLEASE USE: permutations = 1000 or 10000
# BUT NOTE THAT THIS TAKES MUCH MORE COMPUTATION TIME
p.vector <- CombinatorialSignificance(ExpressionSet = PhyloExpressionSetExample,
                                      replicates    = c(2,3,2),
                                      TestStatistic = "FlatLineTest",
                                      permutations  = 100,
                                      parallel      = FALSE)
```

```
 [1] 2.436296e-03 2.288593e-02 1.608399e-03 1.185615e-02 1.835306e-06 1.077012e-05
 [7] 2.025515e-07 5.148342e-07 1.654885e-07 6.251145e-06 9.265520e-10 1.047479e-06
```

As you can see none of the replicate combinations resulted in p-values > 0.05.

```{r,eval = FALSE}
any(p.vector > 0.05)
#> FALSE
```

`CombinatorialSignificance()` can perform all significance tests introduced in the [Introduction](https://github.com/HajkD/myTAI/blob/master/vignettes/Introduction.Rmd) and [Intermediate](https://github.com/HajkD/myTAI/blob/master/vignettes/Intermediate.Rmd) vignettes.

Furthermore, the `parallel` argument allows you to perform significance computations in
parallel on a multicore machine.



