---
title: "Intermediate Phylotranscriptomics Analyses"
author: "Hajk-Georg Drost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intermediate Phylotranscriptomics Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
library(myTAI)
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

In the Introduction vignette you learned how to perform the most fundamental computations
and visualizations for phylotranscriptomics analyses using the `myTAI` package.

This __Intermediate__ vignette will provide you more detailed analyses and more specialized techniques 
to investigate the observed phylotranscriptomics patterns (`TAI`, `TDI`, `RE`, etc.).


## Investigating the statistical significance of phylotranscriptomic patterns

Three methods have been proposed to quantify the statistical significance of the observed phylotranscriptomics patterns
(Quint et al. 2012, Drost et al., 2014).

* __FlatLineTest__
* __ReductiveHourglassTest__
* __EarlyConservationTest__

Here, we will build the test statistic of each test step by step so that future modifications
or new test statistics can be built upon the existing methods implemented in the `myTAI` package.



### FlatLineTest

The FlatLineTest is a permutation test quantifying the statistical significance of a phylotranscriptomic pattern
to deviate from a flat line. The goal is to detect any evolutionary signal within a developmental time course that significantly deviates from a flat line.

To build the test statistic we start with a standard PhyloExpressionSet.
The `myTAI` package provides an example PhyloExpressionSet named `PhyloExpressionSetExample`:

```{r}
library(myTAI)

# load an example PhyloExpressionSet stored in the myTAI package
data(PhyloExpressionSetExample) 

# look at the standardized data set format
head(PhyloExpressionSetExample, 3) 
```

As you can see the first column of the `PhyloExpressionSetExample` stores the Phylostratum assignments of the corresponding genes.
The permutation test is based on random sampling of the Phylostratum assignment of genes. The underlying assumption is that
the `TAI` profile of correctly assigned Phylostrata is significantly deviating from `TAI` profiles based on randomly assigned Phylostrata.

```{r}
# TAI profile of correctly assigned Phylostrata
TAI(PhyloExpressionSetExample) 
```

Visualization: 

```{r, fig.width= 7, fig.height= 5}
# visualize the TAI profile of correctly assigned Phylostrata
PlotPattern(PhyloExpressionSetExample, type = "l", lwd = 6, p.value = FALSE) 
```


```{r}

# randomly permute the Phylostratum assignment of all genes 
randomPhyloExpressionSetExample <- data.frame( sample(PhyloExpressionSetExample[ , 1]), 
                                              PhyloExpressionSetExample[, 2:9] )

# TAI profile based on randomly assigned Phylostrata
TAI(randomPhyloExpressionSetExample) 
```

Visualization: 

```{r, fig.width= 7, fig.height= 5}
# visualize the TAI profile based on randomly assigned Phylostrata
PlotPattern(randomPhyloExpressionSetExample, type = "l", lwd = 6, p.value = FALSE) 
```

As you can see, the visual pattern of the correctly assigned `TAI` profile and the randomly assigned `TAI` profile differ qualitatively.

Now we investigate the variance of the two observed patterns.

```{r}
# variance of the TAI profile based on correctly assigned Phylostrata
var(TAI(PhyloExpressionSetExample)) 
```

```{r}
# variance of the TAI profile based on randomly assigned Phylostrata
var(TAI(randomPhyloExpressionSetExample)) 
```

We observe that the variance of the randomly assigned TAI profile is much smaller than the variance of the correctly assigned TAI profile. Here we use the variance to quantify the _flatness_ of a given TAI profile. In theory the variance of a perfect flat line would be zero. So any TAI profile that is close to zero would resemble a flat line. But how exactly are the variances of randomly assigned TAI profiles distributed. For this purpose the `bootMatrix()` function was implemented.

The `bootMatrix()` takes an PhyloExpressionSet or DivergenceExpressionSet as input and computes $N$ `TAI` or `TDI` profiles of based on randomly assigned Phylostrata or Divergence-Strata.

```{r}
# compute 1000 TAI profiles based on randomly assigned Phylostrata
randomTAIs <- bootMatrix(PhyloExpressionSetExample, permutations = 1000)

head(randomTAIs)

```


Based on this `booMatrix` we can compute the variance of each random TAI profile.



```{r, fig.width= 7, fig.height= 5}
# compute the variance of the random TAI profile for each row
variance_vector <- apply(randomTAIs, 1 , var)

# and visualize the distribution of variances
hist(variance_vector, breaks = 100)
```

Now it is interesting to see where we can find the variance of the correctly assigned TAI.


```{r, fig.width= 7, fig.height= 5}
# variance of the TAI profile based on correctly assigned Phylostrata
var_real <- var(TAI(PhyloExpressionSetExample)) 

# visualize the distribution of variances
hist(c(variance_vector,var_real), breaks = 100, 
     xlab = "variance", main = "Histogram of variance_vector")

# and plot a red line at the position where we can find the 
# real variance
abline(v = var_real, lwd = 5, col = "red")
```

This plot illustrates that variances based on random `TAI` profiles seem have a smaller variance than the variance based on the correct `TAI` profile. To obtain a p-value that now quantifies this difference, we need to fit the histogram of `variance_vector` with a specific probability distribution. 


Visually it would be possible to choose a gamma distribution to fit the histogram of `variance_vector`. To validate this choice a [Cullen and Frey graph](https://openlibrary.org/works/OL1978341W/Probabilistic_techniques_in_exposure_assessment) provided by the [fitdistrplus](http://cran.r-project.org/web/packages/fitdistrplus/index.html) package can be used.


```{r, fig.width= 7, fig.height= 5}
# install.packages("fitdistrplus")

# plot a Cullen and Frey graph
fitdistrplus::descdist(variance_vector)
```

Based on the observation that a gamma distribution is a suitable fit for `variance_vector`, we can now
estimate the parameters of the gamma distribution that fits the data.

```{r, fig.width= 7, fig.height= 5}
# estimate the parameters: shape and rate using 'moment matching estimation'
gamma_MME <- fitdistrplus::fitdist(variance_vector,distr = "gamma",method = "mme")
# estimate shape:
shape <- gamma_MME$estimate[1]
# estimate the rate:
rate <- gamma_MME$estimate[2]

# define an expression written as function as input for the curve() function
gamma_distr <- function(x){ return(dgamma(x = x,shape = shape,rate = rate)) }

# plot the density function and the histogram of variance_vector
curve(gamma_distr,
      xlim = c(min(variance_vector),max(c(variance_vector,var_real))),
      col = "steelblue",lwd = 5,xlab = "Variances", ylab="Frequency")

# plot the histogram of variance_vector
histogram <- hist(variance_vector,prob = TRUE,add = TRUE, breaks = 100)
rug(variance_vector)

# plot a red line at the position where we can find the real variance
abline(v = var_real, lwd = 5, col = "red")
```

Using the gamma distribution with estimated parameters the corresponding p-value of `var_real` can
be computed.

```{r}
# p-value of var_real
pgamma(var_real, shape = shape,rate = rate, lower.tail = FALSE)
```


Hence, the variance of the correct TAI profile significantly deviates from random TAI profiles and this allows us to assume that the underlying TAI profile captures a real evolutionary signal.

This entire procedure of computing the p-value having the variance of TAI profiles as test statistic is done by the `FlatLineTest()` function.

```{r}
# perform the FlatLineTest
FlatLineTest(PhyloExpressionSetExample, permutations = 1000)
```

This function returns the p-value of the test statistic.

