---
title: "Bring your datasets into myTAI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phylo-expression-object}
  %\VignetteEngine{knitr::rmarkdown}
---

## S7 BulkPhyloExpressionSet and ScPhyloExpressionSet objects

Starting from `myTAIv2`, we moved on from the traditional `data.frame` to the new [`S7 OOP system`](https://rconsortium.github.io/S7/) to facilitate fast and flexible computation for evolutionary transcriptomics analyses.

To illustrate this, we will use the example data `example_phyex_set`. The `BulkPhyloExpressionSet` and `ScPhyloExpressionSet` objects are designed to store gene expression data along with associated metadata such as gene ages, phylostratum information, and other relevant properties.

We will first focus on the `BulkPhyloExpressionSet` object, which is for bulk RNA-seq data.

```{r message = FALSE}
library(myTAI)
data("example_phyex_set")
```

```{r message = FALSE, eval=FALSE}
# inspect the object
class(example_phyex_set)
```

<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
class(example_phyex_set)
```
</details>

```{r message = FALSE, eval=FALSE}
S7::prop_names(example_phyex_set)
```
<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
S7::prop_names(example_phyex_set)
```
</details>

```{r message = FALSE, eval = FALSE}
# let's explore the properties using @
# for example:
example_phyex_set@strata |> head()
example_phyex_set@gene_ids |> head()
example_phyex_set@.expression[1:4,1:5]
```

See how rich this `BulkPhyloExpressionSet` object is!

The `BulkPhyloExpressionSet` is designed to ensure efficient interaction with `myTAIv2` functions. For example, if you try the `myTAI::flatline_test` function twice, i.e.

```{r message = FALSE, eval = FALSE}
myTAI::flatline_test(example_phyex_set)
myTAI::flatline_test(example_phyex_set)
``` 

you will notice that the permutations are already pre-computed for the second run so that you don't need to wait a couple of seconds again.

But how do you get your `.tsv` and `.csv` files and convert your `data.frame` and `tibble` to a S7 `BulkPhyloExpressionSet` object?

## Constructing BulkPhyloExpressionSet and ScPhyloExpressionSet

The necessary and sufficient information needed for creating a `BulkPhyloExpressionSet` (as well as a `ScPhyloExpressionSet`) object is:  
(1) A matrix or data frame of gene expression values (genes x samples).  
(2) A numerical value or factored string associated with each gene, typically gene age ranks or [`phylorank`](articles/phylostratigraphy.html) for `TAI` (but also [`deciled dNdS values`](articles/other_strata.html) for `TDI` or [`deciled tau values`](articles/other_strata.html) for `TSI` etc.).

### Loading raw data

The raw outputs of (1) gene expression quantification and (2) gene age inference can typically be found in the form of a tab-separated `.tsv` or comma-separated `.csv` file format. One useful package to do this is `readr`, using the functions `readr::read_csv()` or `readr::read_tsv()`. For `.csv` files, if `readr::read_csv()` doesn't work, try `readr::read_csv2()` in European locales.

### Mock (bulk) dataset for BulkPhyloExpressionSet

#### Bulk RNA-seq data without replicates

In this example, we are using a dataset with only one entry per stage (no replicates).

```{r message = FALSE}
data("example_phyex_set_old")
example_expression <- 
    example_phyex_set_old@.expression |> 
    as.data.frame() |> 
    tibble::rownames_to_column(var = "GeneID")
example_phylorank <-
    example_phyex_set_old@strata
```

```{r message = FALSE, eval = FALSE}
# let's explore the example expression dataset
# for example:
example_expression |> head()
```

<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
example_expression |> head()
```
As you can see, there is one column as the gene identifier and the rest of the columns are the expression values for each developmental stage. Instead of developmental stage, one can also use experimental conditions, e.g. `control`, `treatment`, etc.
</details>
 
Now we construct the input data.frame for `as_BulkPhyloExpressionSet()`

```{r message = FALSE}
example_phyex_set.df <-
    data.frame(phylorank = example_phylorank) |>
    tibble::rownames_to_column(var = "GeneID") |>
    dplyr::select(phylorank, GeneID) |>
    dplyr::left_join(example_expression, by = "GeneID")
```

```{r message = FALSE, eval=FALSE}
example_phyex_set.df |> head()
example_phyex_set.df |> str()
```

<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
example_phyex_set.df |> head()
example_phyex_set.df |> str()
```
</details>

::: {.important}
**These are the requirements for the `as_BulkPhyloExpressionSet()` function:**
column 1 contains the phylorank information, which can be either numeric or factors.  
column 2 contains the gene identifier (here called `GeneID`).  
column 3 onwards contain the expression data with the column titles being the developmental stages or the experimental conditions.
:::

The example `example_phyex_set.df` is now formatted to meet the requirements for the `as_BulkPhyloExpressionSet()` function.

::: {.tip}
The `as_BulkPhyloExpressionSet()` function will automatically detect the first column as the phylorank and the second column as the gene identifier. If your data is structured differently, you can reassign the columns accordingly using `dplyr::relocate()`.
:::

Now, we can use the `as_BulkPhyloExpressionSet()` function to convert the correctly formatted input `data.frame` to a `BulkPhyloExpressionSet` object.

```{r warning=FALSE, message=FALSE}
example_phyex_set.remake <- 
    myTAI::as_BulkPhyloExpressionSet(
        example_phyex_set.df,
        name = "reconstituted phyex_set_old")
```

There you have it, a `BulkPhyloExpressionSet` object!

```{r message = FALSE, eval=FALSE}
# inspect the object
class(example_phyex_set.remake)
```

<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
class(example_phyex_set.remake)
```
</details>

```{r message = FALSE, eval=FALSE}
S7::prop_names(example_phyex_set.remake)
```

<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
S7::prop_names(example_phyex_set.remake)
```
</details>

You can further explore the properties of the `BulkPhyloExpressionSet` object using the `@` operator, for example: `example_phyex_set.remake@counts |> head()`.

... and plot your transcriptome age index!

```{r message = FALSE, fig.height=4, fig.width=6, fig.alt="plot_signature_output", dev.args = list(bg = 'transparent'), fig.align='center'}
example_phyex_set.remake |> 
    myTAI::plot_signature()
```


#### What if the dataset contains replicates?

This is exactly the case for the new `example_phyex_set` dataset (not like `example_phyex_set_old`), which contains replicates for each developmental stage. The `as_BulkPhyloExpressionSet()` function can handle this as well, as long as we use the `groups` parameter to specify the grouping of replicates, i.e.

```{r message = FALSE}
data("example_phyex_set") # not data("example_phyex_set_old")
example_expression <- 
    example_phyex_set@.expression |> 
    as.data.frame() |> 
    tibble::rownames_to_column(var = "GeneID")
example_phylorank <-
    example_phyex_set@strata

example_phyex_set.df <-
    data.frame(phylorank = example_phylorank) |>
    tibble::rownames_to_column(var = "GeneID") |>
    dplyr::select(phylorank, GeneID) |>
    dplyr::left_join(example_expression, by = "GeneID")
```

```{r message = FALSE, eval=FALSE}
colnames(example_phyex_set.df)
```
<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
colnames(example_phyex_set.df)
```
</details>

As you can see, this dataset has three replicates per stage.

```{r message = FALSE}
groups_example_phyex_set.df <- c(
            rep("Preglobular",3), 
            rep("Globular",3), 
            rep("Early Heart",3), 
            rep("Late Heart",3), 
            rep("Early Torpedo",3), 
            rep("Late Torpedo",3), 
            rep("Bent Cotyledon",3), 
            rep("Mature Green",3))

# we can add the group information using groups
example_phyex_set.remake <- 
    myTAI::as_BulkPhyloExpressionSet(
        example_phyex_set.df,
        groups = groups_example_phyex_set.df, # adding group information here
        name = "reconstituted phyex_set")
```

Done!

::: {.tip}
Along with the `groups` parameter, you can also specify the `name`, `strata_labels` and other parameters to provide a more descriptions for your `BulkPhyloExpressionSet` object. This is useful for plotting and visualisation purposes. Check this using `?` before the function (i.e. `?myTAI::as_BulkPhyloExpressionSet()`.
:::

### Mock single-cell dataset for ScPhyloExpressionSet

If you are interested in single cell RNA-seq data, the process is similar, but you will need to ensure that your data is structured correctly for single cell analysis, i.e. as a [`seurat object`](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html). 

You can then use the `as_ScPhyloExpressionSet()` function with adjustments to the relevant metadata.

```{r message = FALSE, warning=FALSE}
# Load the relevant packages
library(dplyr)
library(Seurat)

pbmc_raw <- read.table(
  file = system.file('extdata', 'pbmc_raw.txt', package = 'Seurat'),
  as.is = TRUE
)
```

The input file for `Seurat::CreateSeuratObject()` is a tab-separated file with the first column as gene identifiers and the rest of the columns as expression values for each cell. The `pbmc_raw` object is a data frame with gene expression data.

```{r message = FALSE, eval = FALSE}
head(pbmc_raw)
```
<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
head(pbmc_raw)
```
</details>

Now, we create a Seurat object. i.e.

```{r message = FALSE, eval = FALSE}
pbmc_small <- CreateSeuratObject(counts = pbmc_raw)
```

```{r message = FALSE, eval = FALSE}
is(pbmc_small)
```
<details> <summary>Show output</summary>
```{r message = FALSE, echo=FALSE}
is(pbmc_small)
```
</details>

Now, we can convert this Seurat object to a `ScPhyloExpressionSet` object using the `as_ScPhyloExpressionSet()` function.

The next part is a bit tricky, as you need to provide the phylorank information in a specific format with the same gene identifiers as that of `pbmc_small`.

We will generate **random** phyloranks for each gene as integers.

```{r message = FALSE}
gene_names <- pbmc_small |> rownames()
example_phylorank_sc <- 
    setNames(sample(1:10, length(gene_names), replace = TRUE), gene_names) |>
    as.factor()
```

```{r message = FALSE}
example_phyex_set_sc <- 
    myTAI::as_ScPhyloExpressionSet(
        pbmc_small,
        strata = example_phylorank_sc)
```

<details> <summary>Example of a downstream analysis</summary>
Here, we will go a bit deeper and cluster the `pbmc_small` object before integrating this to the `myTAIv2` workflow.

```{r message = FALSE, warning=FALSE}
# example workflow to cluster the single cell data
pbmc_small <- Seurat::NormalizeData(pbmc_small)
pbmc_small <- Seurat::FindVariableFeatures(pbmc_small, selection.method = "vst", nfeatures = 20)
pbmc_small <- Seurat::ScaleData(pbmc_small)
pbmc_small <- Seurat::RunPCA(pbmc_small, features = VariableFeatures(object = pbmc_small))
pbmc_small <- Seurat::FindNeighbors(pbmc_small, dims = 1:10)
pbmc_small.cluster <- Seurat::FindClusters(pbmc_small, resolution = 0.8)
```

```{r message = FALSE}
example_phyex_set_sc.cluster <- 
    myTAI::as_ScPhyloExpressionSet(
        pbmc_small.cluster,
        strata = example_phylorank_sc)
```

Now we have the `ScPhyloExpressionSet` object, which you can now plot!

```{r message = FALSE, fig.height=4, fig.width=6, fig.alt="plot_signature_output", dev.args = list(bg = 'transparent'), fig.align='center'}
plot_signature(example_phyex_set_sc.cluster)
```

::: {.important}
This example used an example Seurat object with a mock phylorank dataset (random gene age assignment) and an example workflow. In practice, you will have to follow the best standards for single cell RNA-seq data analysis, including quality control, normalisation, and clustering steps. And assign real phylorank information to the genes in your dataset (to do this, see the [phylostratigraphy vignette](articles/phylostratigraphy.html)).
:::
</details>

## Summary

In this section, we have learned how to create `BulkPhyloExpressionSet` and `ScPhyloExpressionSet` objects from raw gene expression data and phylorank information. This object is essential for performing evolutionary transcriptomics analyses using the `myTAI` package, and provides a structured way to store and manipulate evolutionary transcriptomics data, making it easier to perform various analyses and visualisations.
